---
title: "Data Discovery via API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Discovery via API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Global Search allows users to search for vaults, files, folders, and datasets by name, tags, user, date, and other metadata that can be customized. Similarly to Global Search on the web application, the search functionality is available through EDP Python and R clients as well.

## Global Search Basics

The EDP Global Search performs a search based on the provided set of parameters (filters, entities, query, limit, ordering, etc.):

-   query: Advanced search query string
-   filters: Filters to apply
-   entities: List of entity tuples to filter on (entity type, entity)
-   limit: Maximum number of query results to return

```{r}
library("quartzbio.edp")

# By default it limits the number of objects in search result to 100
results <- GlobalSearch.search()

# To return all objects set parameter paginate to TRUE
results <- GlobalSearch.search(paginate = TRUE)
```

Users may use the limit parameter to limit the number of returned objects:

```         
# No filters applied with limit parameter
results <- GlobalSearch.search(limit = 200)
```

## Applying filters for Global Search

Similar to the web application, users can apply filters with R client:

```{r}
library("quartzbio.edp")

# Searching only for vaults
response <- GlobalSearch.search(filters = '[{"and":[["type__in",["vault"]]]}]')

# Searching based on date created
response <- GlobalSearch.search(filters = '[{"and":[{"and":[["created_at__range",["2023-01-01","2023-12-31"]]]}]}]')
```

## Advanced Search Query

Users can write their own queries the using R client by providing query parameters:

```{r}
library("quartzbio.edp")

# Advanced search (using keyword argument)
results <- GlobalSearch.search(query = "TCGA", paginate = TRUE)

# Advanced search (using positional argument)
results <- GlobalSearch.search("test")
```

## Global Beacon Search

Global Beacon Search can be performed as well with both Python and R clients by using the entities parameter. Please note that Global Beacon Search works only on datasets enabled by Global Beacons. To search for subjects or samples, users should also set the vault_scope parameter to "any".

```{r}
# Entity search example
GlobalSearch.search(entities = '[["gene","BRCA2"]]')

# Entity search example
GlobalSearch.search(entities = '[["variant", "GRCH38-7-140753336-140753336-T"]]')

# Sample entity search example
GlobalSearch.search(entities = '[["sample","A00001"]]', vault_scope = "any")
```

## Retrieving Subjects with Global Search

```{r}
# Getting the subjects
GlobalSearch.subjects(entities = '[["gene","BRCA2"]]')
```
