---
title: "EDP Use Cases with R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDP Use Cases with R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

In this article, you can find examples of using EDP with the R client.

## Global Beacons

Global Beacon lets anyone in your organization find datasets based on the entities it contains (i.e. variants, genets, targets). Only datasets that have entities can be indexed.

For more information about Global Beacons, please review the Global Beacons Overview article.

```{r}
library("quartzbio.edp")
connect(secret = "TOKEN")
```

First, let's start with enabling Global Beacon on the dataset:

```{r}
# Dataset ID
dataset_id <- "1658666726768179211"


# Turn on Global Beacon on the selected dataset
Object.enable_global_beacon(dataset_id)
```

Let's check now the status of Global Beacon indexing for the datasets:

```{r}
# Waiting a minute until indexing is complete
Sys.sleep(60)

# Getting the status of global beacon on the dataset
Object.get_global_beacon_status(dataset_id)
```

As we can see from the response, the indexing is complete. It means that now we can perform a search on dataset entities.

Similarly to enabling Global Beacon for the datasets, you may also disable it using the disable_global_beacon function:

```{r}
Object.disable_global_beacon("1676139881237207342")
```

## Global Search

Global Search allows you to search for vaults, files, folders, and datasets by name, tags, user, date, and other metadata which can be customized.

For more information about Global Search, please review the Global Search Overview article.

Similarly to Global Search on the web application, the search functionality is available through the EDP R client as well.

### Global Search functions

GlobalSearch module provides three functions and all of them use the same set of parameters (filters, entities, query, limit, offset, etc):

**1. GlobalSearch.search**

Performs a global search based on provided filters, entities, and queries (advanced query) and returns an R data frame containing results from API response. For a full list of results set paginate attribute to TRUE:

```{r}
GlobalSearch.search(filters = '[{"and":[["type__in",["dataset"]]]}]')
GlobalSearch.search(query = "fuji")
GlobalSearch.search(query = "fuji", paginate = TRUE)
```

**2. GlobalSearch.subjects**

Similar to the search function, the subjects function returns the results in the form of an R data frame. The returned data frame contains subjects:

```{r}
GlobalSearch.subjects(entities = '[["gene","BRCA2"]]')
```

**3. GlobalSearch.request**

Performs low-level global search based on the provided filters, similar to what you would do from a web application. In the response there will be the following attributes:

-   results - list of vault objects (datasets, files, folders, and vaults). Those are the same - - objects from the "Results" tab on the GlobalSearch page on Mesh.
-   total - number of objects in the search results
-   vaults - list of vaults
-   subjects - list of subjects
-   subjects_count - number of subjects in the subjects list
-   took - the amount of time it took to perform a search
-   offset - offset for pagination, offset from the first result you want to fetch

You may call the request function by providing some of the following arguments: filters, entities, query (advanced search query), limit, and offset:

```{r}
GlobalSearch.request(query = "fuji", limit = 200)
GlobalSearch.request(entities = '[["gene","BRCA2"]]')
GlobalSearch.request(entities = '[["gene","BRCA2"]]', filters = '[{"and":[{"and":[["created_at__range",["2021-11-28","2021-12-28"]]]},["type__in",["dataset"]]]}]')
```

**Recommended functions to use:**

-   GlobalSearch.search - for getting the search results.
-   GlobalSearch.subjects - for getting the subjects.

### Search Examples

1.  Global Beacon Search

As we previously indexed the dataset, we should be able to perform an entity search and see that dataset in the results.

```{r}
results <- GlobalSearch.search(entities = '[["gene","BRCA2"]]')
results
```

Each result object has the following attributes:

```{r}
names(results)
```

2.  Applying filters for Global Search

Search only for vaults:

```{r}
response <- GlobalSearch.search(filters = '[{"and":[["type__in",["vault"]]]}]')
response
```

Search based on the date created:

```{r}
response <- GlobalSearch.search(filters = '[{"and":[{"and":[["created_at__range",["2021-11-21","2021-12-28"]]]}]}]')
response
```

3.  Advanced search query

The search function has embedded pagination in itself, so by setting that attribute to true, it will fetch all results (warning: that operation may be costly and time-consuming).

By performing an advanced search using query argument, only 100 objects are returned and you may see the reason for that in the following output message:

```{r}
# Advanced search
response <- GlobalSearch.search(query = "fuji")

# Number of objects in the response
dim(response)
```

We can perform a request function call to get the full API response and see how many results we have in total:

```{r}
response <- GlobalSearch.request(query = "fuji")
response$total
```

There are 1407 objects in total but we have only 100 of them returned as default. In order to get all the results you may use parameter paginate = TRUE (please note that retrieving all objects may take a while):

```{r}
results <- GlobalSearch.search(query = "fuji", paginate = TRUE)
print(dim(results))
```

Alternatively, instead of using paginate parameter, you may use the limit parameter instead. Here we're setting a limit to 500 objects:

```{r}
results <- GlobalSearch.search(query = "fuji", limit = 500)
print(dim(results))
```

4.  Getting the Global Search subjects

Similar to the search function to get the result objects in the previous sections, we can use the subjects function to get a data frame containing only subjects:

```{r}
GlobalSearch.subjects(entities = '[["gene","BRCA2"]]')
```

### Annotate with Dataset Queries

EDP makes it possible to annotate a field with data from any other dataset. Users can run any dataset query in an expression and use it to annotate a field.

In this example, we have a list of variants in the EDP variant ID format. We will annotate them with information about dataset presence (where has this variant been seen?), clinical significance, population allele frequencies, and prevalence in cancer types. Users can easily add their own annotation datasets by importing them into EDP and querying them in an expression.
