---
title: "Connection to EDP hosts"
output:
  rmarkdown::html_vignette:
    toc: true
description: |
  describe connections methods available
vignette: >
  %\VignetteIndexEntry{Connection to EDP hosts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, echo = FALSE, message =  FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
withr:::defer({Sys.unsetenv('EDP_PROFILE')}, parent.frame(n=2))
```

# TL;DR

Save into the  **~/.qb/edp.json**  files, different EDP configuration profiles.  
The **default** profile is used automatically in functions if no connection are specified in arguments.

```
{
  "default": {
    "secret": "API_TOKEN",
    "host": "https://api.solvebio.com"
  },
  "demo-corp":  {
    "secret": "API_TOKEN_2",
    "host": "https://demo-corp.api.solvebio.com"
  }
}
```

```{r}

# by default all functions will use the default profile set in the default config file
library(quartzbio.edp)
User()
vlst <- Vaults()
vlst[[1]]

flds <- Folders()
flds[[1]]

fis  <- Files()
fis[[1]]

das  <-Datasets()
das[[1]]

```

# Introduction

There are several methods to be connected to a EDP instance.  

* using explicit connection
* using environment variables
* using profiles
* using autoconnect()

* **key**: designates an API key (deprecated and will not be supported anymore in the future)
* **token**: designates an API token
* **secret**: can be a *key* or a *token* but *token* connections based are highly recommended 


## Connection from Profiles

Different EDP environment profiles config can be stored in the default config location of your home 
directory: **~/.qb/edp.json** 
At least a single **default** profile should be defined as examplified above.  

* the default profile witl set the default connection, if no explicit connection was specified using
the environment variables.


```{r}

# use the default profile
User()

# use explicitely the default profile
conn <- connect_with_profile()
User(conn)

# use  profile named demo-corp defined in the default config file without
# changing the current connection
conn <- connect_with_profile('demo-corp')
User(conn)
Vaults(conn = conn)

# still using the default profile
# because no explicit profile connection is given in argument
Vaults()

```

## Connection From System Environment variables

If no **default** profile or config are available, the environment variables **EDP_PROFILE** and **EDP_CONFIG** are explored to obtain the profile name.  
This way several config files, containing different profiles can be used by switching only environment variables.  

Setting *EDP_PROFILE to 'default'* and *EDP_CONFIG to '~/.qb/edp.json'* is equivalent to use the default
package settings.  

A good practice is to have profiles defined in the **EDP_CONFIG** file and to set the **EDP_PROFILE** environment variable at
the beginning of your scripts.  The remaining connections will automatically use it.

```{r, include = FALSE, echo = FALSE, message =  FALSE, warning=FALSE}
  # usnet connection for next block
  set_connection(NULL)
```

```{r}
# set the demo-corp connection as the default for future connections

Sys.getenv('EDP_PROFILE')
Sys.setenv(EDP_PROFILE = 'demo-corp')
User()

```

## Explicit connection

**secret** and **host** can be set explicitely. *read_connection* returns connection infos for a 
given config.

```{r}
conn <- read_connection_profile('default')
secret <- conn$secret
host <- conn$host
conn <- connect(secret, host)
User(conn)

```
## autoconnect()

**autoconnect()** function tries to connect using the environment variables **EDP_PROFILE** and 
**EDP_CONFIG** at first and  then the **default** profile in the default config file.  

```{r}
# auto connect use the EDP profile to set the current connection
Sys.setenv(EDP_PROFILE = 'demo-corp')
conn <- autoconnect()
User(conn = conn)
User()


# unset the EDP_PROFILE environment variable
# autoconnect() use the default settings.
Sys.unsetenv('EDP_PROFILE') # unset current profile environment variables
set_connection(NULL) # unset current connection

conn <- autoconnect()

# Will used the default profile
User(conn=conn)
User()

```

# Advanced Usage 

## set_connection() and get_connection()

So far connection objects (i.e list) were passed as arguments to functions such as *User()* or *Vaults()*.  
Still theses functions can use a current connection.   
By default, if no connection argument is given to these function(). The *current connection* will 
be set to the default connection. 

* **get_connection()** returns the current connection. If none are yet available and **auto=TRUE**, it 
will use autoconnect() to obtain it and sets as the current one.

* **set_connection()**: explicitely sets the *current connection*


```{r}

# explicitely set the connection with a profile
set_connection( connect_with_profile('demo-corp'))

# remove the current connection
set_connection(NULL)

# the firt call to User() sets the current default connection to the default connection
# both calls will use the current connection

User()
User(conn = get_connection())

# the current connection remains the 'default'
# connecting with a profile did not change it
conn_corp <- connect_with_profile('demo-corp')
User(conn_corp)
User()

# the current connection will be the 'demo-corp' one 
# after unsetting the previous connection and profile environment variables
Sys.setenv(EDP_PROFILE = 'demo-corp')
set_connection(NULL)
User()

```
